---
title: "Reproduceable Document"
subtitle: "How to make document that can be easily updated"

author: Max Müsterman, Fredericke Kruger


output:
    
  word_document:
    toc: true # table of contents
    toc_depth: 3 # number of levels for table of contents
    highlight: tango # color formatting for code chunks
    reference_docx: template.docx # for setting up basic style of Word doc

papersize: a4
bibliography: refs.bib # bibtext file containing bibliographic records 
csl: style.csl # style for reference formating (can be downloaded for specific journal)
---

```{r setup, include=FALSE}
library(knitr) # for Rmarkdown

library(png) # for reading in pngs
library(formatR) # for making model summary tables

library(flextable) # for lots of table options
library(kableExtra) # extend the basic functionality of tables produced using knitr 


knitr::opts_chunk$set(
  cache = FALSE, # TRUE for speed. Remember to empty cache after changes to R code sections
  cache.path = 'cache_supplMat/', fig.path  ='tex_supplMat/',
  # fig.align='center',
  comment = NA,
  message = FALSE, warning = FALSE, echo = TRUE,
  tidy.opts=list(width.cutoff = 60), tidy = TRUE # automatically wraps code
)

options(knitr.kable.NA = '')

# set global flextable layout
fonttype <- "Arial"
set_flextable_defaults(
  # font.family = "Palatino Linotype",
  # hansi.family = "Palatino Linotype",
  # cs.family = "Palatino Linotype",
  font.family = fonttype,
  hansi.family = fonttype,
  cs.family = fonttype,
  font.size = 8.5,
  theme_fun = "theme_vanilla",
  big.mark = "" # no 1000 separator
)


# set.seed(1)

# counters for labelling figures and tables 
## I prefer this so that I can make internal citations that are dynamic
iFig = 0
iTab = 0
```

```{r makeOutput, warning=FALSE, include=FALSE}
# make some data and plots
tmpDir <- tempdir()

# plots
fname <- file.path(tmpDir, "volcano.png")
png(fname, width = 3, height = 2, units = "in", res = 400)
  op <- par(mar = c(3,3,1,1), ps = 8)
  image(volcano)
  par(op)
dev.off()

n <- 1e5
dat <- data.frame(x = rnorm(n), y = rnorm(n))
fname <- file.path(tmpDir, "points.png")
png(fname, width = 5, height = 5, units = "in", res = 400)
  op <- par(mar = c(4,4,1,1))
  plot(y ~ x, dat, pch = ".", cex = 3, col = adjustcolor(1,0.1))
  par(op)
dev.off()

P <- prcomp(as.matrix(iris[,1:4]))
iris2 <- cbind(iris, P$x[,1:2])

fname <- file.path(tmpDir, "iris_pca.png")
png(fname, width = 5, height = 5, units = "in", res = 400)
  op <- par(mar = c(4,4,2,1))
  plot(PC2~PC1, iris2, col = Species, pch = 16)
  legend("topright", legend = unique(iris2$Species), col = seq(unique(iris2$Species)), pch = 16)
  mtext("PCA (iris morphology)", adj = 0)
  par(op)
dev.off()

# tables
fname <- file.path(tmpDir, "iris2.Rdata")
save(iris2, file = fname)

gsub("\\\\", "/", tmpDir) # to get path

```

\pagebreak

# Basic outputs from R *chunks*

## Run code and output directly

This can be useful for simple outputs, but it is a bit harder to control plot formatting in my opinion. By setting "echo = FALSE", the call to R is not shown - only the output:  

```{r echo=FALSE, fig.cap="Figure X. Caption information.", fig.dpi=400}
plot(iris)
```

<br>

<br>

\pagebreak

## Change the figure dimensions  

Note that '"fig.align = "center"' doesn't work in a Word document.  

```{r echo=FALSE, fig.dpi=400, fig.height=4, fig.width=4}
par(ps = 8)
plot(iris)
```

# Citations

-   Citation with author "[\@pastoors_effects_2000]": [@pastoors_effects_2000]
-   Multiple citations "[\@chong_performance_2020; \@taylor_using_2021]": [@chong_performance_2020; @taylor_using_2021]
-   Cited in line "\@friedland_trends_2021": @friedland_trends_2021
-   Citation without author "[-\@chong_performance_2020]": [-@chong_performance_2020]

These automatically get added to a list of references at the end. Thus, a final header for "References" should be added.  

# Figures

In the hidden code chunk at the top, I have added counters for figures and tables, (iTab, iFig). To print out the counter, you will need to perform an in-line call to R using back ticks, \`r\`, to both set it to the next value and print it. For example, my next figure will be Figure `r iFig+1`. I included the code "r iFig+1", which printed the next counter value. I usually do all the setting of new counter values in the caption of the Figure itself; e.g. the caption below contains the inline R code "r iFig=iFig+1; iFig", which first advances the counter and then prints the value.  

```{r echo=FALSE, fig.dpi=400}
par(mar = c(4,4,1,1))
image(volcano)
```

**Figure `r iFig=iFig+1; iFig`.** This figure shows bla bla.  

Finally, for best control on figure output, I usually try to work with plots that have been previously rendered according to the specifications that I want. A png format is usually a good option due to file size and quality (output with knitr::include_graphics). Again, I have advanced the figure counter, and Figure `r iFig+1` should be approximately 2 by 3 inches in size.  

```{r echo=FALSE}
fname <- file.path(tmpDir, "volcano.png")
knitr::include_graphics(fname)
```

**Figure `r iFig=iFig+1; iFig`.** This figure shows the elevation of a volcano.  

<br>

<br>

\pagebreak

And one last example, Figure `r iFig+1` should be approximately 5 by 5 inches in size.

```{r echo=FALSE, fig.dpi=400}
fname <- file.path(tmpDir, "points.png")
knitr::include_graphics(fname)
```

**Figure `r iFig=iFig+1; iFig`.** A plot of a bunch of random points.
  
  
  

\pagebreak

# Tables

## Standard output

<br>

Using the *flextable* package, we can control the look of tables much better (e.g. output of *iris* dataset, Table `r iTab+1`). See [flextable book](https://ardata-fr.github.io/flextable-book/) for lots of examples.  



**Table `r iTab=iTab+1; iTab`.** Head of *iris* dataset.  

```{r echo=FALSE}
fname <- file.path(tmpDir, "iris2.Rdata")
load(file = fname)

ft <- flextable(head(iris2, 10))
ft <- bg(ft, part = "header", bg = "grey80")
ft <- set_table_properties(ft, layout = "autofit")
ft
```

<br> <br>

\pagebreak

## Model output

First, we see the direct printing of a linear model summary:  

<br>

<br>

  

```{r}
fit <- lm(Petal.Length ~ Petal.Width*Species, data = iris)
summary(fit)
```

<br>

<br>

\pagebreak

Then, we see a better version using the *flextable* package (Table `r iTab+1`):

<br>

**Table `r iTab=iTab+1; iTab`.** Clean model summary from *as_flextable*.

```{r echo=FALSE}
ft <- as_flextable(fit)

ft <- bg(ft, part = "header", bg = "grey80")
ft <- set_table_properties(ft, layout = "autofit")
ft

```

\pagebreak

# References
