---
title: "Supplementary materials"
subtitle: "Figures and Tables"

author: 
  - Max Müsterman
  - Freddy Kruger


output:
  
  pdf_document:
    latex_engine: xelatex
    number_sections: TRUE
    toc_depth: 3
    highlight: tango
    keep_tex:  true
    
  # word_document:
  #   # toc: true # inclusion will prevent update when .doc is open
  #   toc_depth: 3
  #   highlight: tango
  #   reference_docx: template.docx

papersize: a4
bibliography: refs.bib
csl: style.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE, # TRUE for speed. Remember to empty cache after changes to R code sections
  cache.path = 'cache_supplMat/', fig.path  ='tex_supplMat/',
  fig.align='center',
  comment=NA,
  message=FALSE, warning=FALSE, echo=TRUE,
  tidy.opts=list(width.cutoff=60), tidy=TRUE # automatically wraps code
)

options(knitr.kable.NA = '')


library(png)
library(formatR)
library(knitr)


# set.seed(1)
iFig = 0
iTab = 0
```

\pagebreak

# Introduction

Bla di bla [@pastoors_effects_2000]. @friedland_trends_2021 results support other studies [@chong_performance_2020;  @taylor_using_2021]. Cite Pastoors without adding name [-@pastoors_effects_2000]

# Methods

```{r, echo=FALSE}
fname <- file.path("output/volcano.png")
tmp <- readPNG(fname)
knitr::include_graphics(fname, dpi = floor(dim(tmp)[2]/6))
```

**Figure S`r iFig=iFig+1; iFig`.** bla di bla bla.

```{r, echo=FALSE}
fname <- file.path("output/mvrnorm.png")
tmp <- readPNG(fname)
knitr::include_graphics(fname, dpi = floor(dim(tmp)[2]/6))
```

**Figure S`r iFig=iFig+1; iFig`.** bla bla




```{r}
library(vegan)
data(dune)
data(dune.env)
mod <- cca(dune ~ A1 + Moisture + Management, dune.env)
## better control -- remember to set scaling etc identically
plot(mod, type="n", scaling="sites")
text(mod, dis="cn", scaling="sites")
points(mod, pch=21, col="red", bg="yellow", cex=1.2, scaling="sites")
text(mod, "species", col="blue", cex=0.8, scaling="sites")
```

**Figure S`r iFig=iFig+1; iFig`.** CCA plot example.


## subsection


### sub-subsection


\pagebreak

# Results

## subsection

**Table S`r iTab=iTab+1; iTab`.** bla bla

```{r, echo = FALSE}
load("output/iris3.Rdata")
df <- as.data.frame(iris3[1:8,])
knitr::kable(df, digits = 3)
# kableExtra::kbl(df, digits = 3, booktabs = T, linesep = "") # useful for pdfs and html formatting
```

Notes: Some *notes*. \*All variables showed significant effects at the p ≤ 0.001 level.

\pagebreak

```{r, echo = FALSE}
fname <- file.path("output/volcano.png")
tmp <- readPNG(fname)
knitr::include_graphics(fname, dpi = floor(dim(tmp)[2]/6))
```

**Figure S`r iFig=iFig+1; iFig`.** bla di bla bla.

\pagebreak

# References